Group Stage Play

### 2025-10-10 group_stage.qmd
CMSAC presentation

```{r}
simulate_group_play <- function(strengths, rounds=1){
  teams <- strengths$true_rank
  n <- length(teams)
  win_counts <- numeric(n)
  loss_counts <- numeric(n)
  for (i in 1:rounds){
    for (i in seq_along(teams)){
      for (j in seq_along(teams)){
        if (i != j){
          team1 <- teams[i]
          team2 <- teams[j]
          strength1 <- strengths$true_strength[strengths$true_rank == team1]
          strength2 <- strengths$true_strength[strengths$true_rank == team2]
          match_winner <- simulate_match(teams[i],teams[j],strength1, strength2)
          if (match_winner == teams[i]) {
            win_counts[i] <- win_counts[i] + 1
            loss_counts[j] <- loss_counts[j] + 1
          } else {
            win_counts[j] <- win_counts[j] + 1
            loss_counts[i] <- loss_counts[i] + 1
          }
        }
      }
    }
  }
  results <- data.frame(
    true_rank = teams,
    game_wins = win_counts,
    game_loses = loss_counts,
    true_strength = strengths$true_strength,
    groups = strengths$groups
  )

  # Ranks teams based on number of wins (includes ties)
  results$rank_hat <- rank(-results$game_wins, ties.method = "random")
  results <- results[order(results$rank_hat),]

  return(results)
}

simulate_groups <- function(group_length, num_teams, distribution, num_advance){
  teams <- paste(num_teams:1)
  if (distribution == "Normal"){
    strengths <- sapply(num_teams, function(n) { qnorm(1:n/(n+1)) })
    normal_strengths <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      rank_hat = rep(NA,length(teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams))
    )
    df <- arrange(normal_strengths, true_rank)
  }
  else if (distribution == "Same"){
    strengths <- sapply(num_teams, function(n) {rep(0,n)})
    normal_strengths <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      rank_hat = rep(NA,length(teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams))
    )
    df <- arrange(normal_strengths, true_rank)
  }
  else if (distribution == "Uniform"){
    strengths <- sapply(num_teams, function(n) { qunif(1:n/(n+1) , 0 , sqrt(12)) })
    unif_strength <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams)),
      rank_hat = rep(NA,length(teams))
    )
    df <- arrange(unif_strength, true_rank)
  }

  num_groups <- num_teams / group_length
  # Create tiers based on sorted true_strength
  df <- df %>% arrange(desc(true_strength))
  tiers <- split(df, ceiling(seq_along(df$true_rank) / num_groups))

  # Initialize group vector
  df$groups <- NA

  # Assign one team from each tier to each group randomly
  for (tier in tiers) {
    group_ids <- sample(1:num_groups)  # random group order for this tier
    df$groups[match(tier$true_rank, df$true_rank)] <- group_ids
  }

  all_results <- list()
  for (g in unique(df$groups)){
    group_df <- df[df$groups == g, ]
    group_results <- simulate_group_play(strengths = group_df)
    group_results$groups <- g
    all_results[[g]] <- group_results
  }
  print(all_results)
  final_df <- do.call(rbind, all_results)
  teams_in_knockout <- final_df %>%
    arrange(rank_hat, game_wins) %>%
    head(num_advance * num_groups)
  return(teams_in_knockout)
}
create_knockout_matches <- function(df) {
  group_list <- sort(unique(df$groups))
  matches <- list()

  for (i in seq(1, length(group_list), by = 2)) {
    g1 <- group_list[i]
    g2 <- group_list[i + 1]

    # Get the winner from group g1 and runner-up from group g2
    team1 <- df[df$groups == g1 & df$rank_hat == 1, ]
    team2 <- df[df$groups == g2 & df$rank_hat == 2, ]

    # Get the winner from group g2 and runner-up from group g1
    team3 <- df[df$groups == g2 & df$rank_hat == 1, ]
    team4 <- df[df$groups == g1 & df$rank_hat == 2, ]

    matches[[length(matches) + 1]] <- data.frame(
      Match = paste("Match", length(matches) + 1),
      TeamA = team1$true_rank,
      TeamB = team2$true_rank
    )

    matches[[length(matches) + 1]] <- data.frame(
      Match = paste("Match", length(matches) + 1),
      TeamA = team3$true_rank,
      TeamB = team4$true_rank
    )
  }

  do.call(rbind, matches)
}
simulate_knockout <- function(df, series=1, third_place = T){

  seeding_structure <- create_knockout_matches(df)

  for (i in nrow(df)){
    for (g in unique(df$groups)){

    }
  }

  results <- list()  # Initialize the results list

  for (i in 1:nrow(seeding_structure)) {
    # Create a new data frame based on the current permutation in 'test'
    permutation <- seeding_structure[i, ]
    permuted_df <- df
    permuted_df$true_rank <- df$true_rank[permutation]  # Update team order 
    teams <- permuted_df$true_rank
    permuted_df$true_strength <- df$true_strength[permutation]
    permuted_df$rank_hat <- NA  # Initialize ranks column
    permuted_df$game_wins <- 0  # Initialize wins column
    permuted_df$game_losses <- 0

    strengths <- permuted_df$true_strength
    round_number <- ceiling(log2(length(teams)))

    # Track the losers of the semifinals for the third-place match
    losers_semis <- c()

    while (length(teams) > 1) {
      #cat("\n--- New Round ---\n")
      next_round <- c() # Makes sure only winners are included in next round

      for (j in seq(1, length(teams), by = 2)) { # Gets 2 teams for each match
        team1 <- teams[j]
        team2 <- teams[j+1]

        strength1 <- strengths[which(permuted_df$true_rank == team1)]
        strength2 <- strengths[which(permuted_df$true_rank == team2)]

        wins_team1 <- 0
        wins_team2 <- 0
        losses_team1 <- 0
        losses_team2 <- 0
        for (game in 1:series){
          match_winner <- simulate_match(team1, team2, strength1, strength2)
          #cat(sprintf("Matchup:  %s vs %s | Strengths: %.2f vs %.2f | Winner: %s\n", team1, team2, strength1, strength2, match_winner))
          if(!is.na(team1)) {
          if (match_winner == team1) {
            wins_team1 <- wins_team1 + 1
            losses_team2 <- losses_team2 + 1
          } else {
            wins_team2 <- wins_team2 + 1
            losses_team1 <- losses_team1 + 1
          }
          }
          else {wins_team2 <- series / 2 + 1}

          # Determine the overall winner if a majority is reached
          if (wins_team1 > series / 2 || wins_team2 > series / 2) {
            break
          }
        }

        # adds wins and losses of the series to the teams unless they have a bye
        if (!any(is.na(c(team1, team2)))) {
          permuted_df <- permuted_df %>%
            mutate(game_wins = ifelse(true_rank == team1, game_wins + wins_team1, game_wins)) %>%
            mutate(game_wins = ifelse(true_rank == team2, game_wins + wins_team2, game_wins)) %>%
            mutate(game_losses = ifelse(true_rank == team1, game_losses + losses_team1, game_losses)) %>%
            mutate(game_losses = ifelse(true_rank == team2, game_losses + losses_team2, game_losses)) 
           }

        # Determine the match winner based on series results
        series_winner <- ifelse(wins_team1 > wins_team2, team1, team2)
        #cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, series_winner))
        next_round <- c(next_round, series_winner)

        if (round_number > 2){
          # Assign ranks based on the current round
          n <- 2^(round_number)
          #print(series_winner)
            if (!is.na(team1) & series_winner == team1) {
                permuted_df$rank_hat[permuted_df$true_rank == team2] <- n
            } else {
                permuted_df$rank_hat[permuted_df$true_rank == team1] <- n
            }
        } else if (round_number == 2) {  # Semifinals
          n <- 2^(round_number)
          if (match_winner == team1) {
              permuted_df$rank_hat[permuted_df$true_rank == team2] <- n
            losers_semis <- c(losers_semis, team2)
          } else {
              permuted_df$rank_hat[permuted_df$true_rank == team1] <- n
            losers_semis <- c(losers_semis, team1)
          }
        }else if (round_number == 1) { # Finals
          permuted_df <- permuted_df %>%
            filter(!is.na(true_rank))
          n <- 2^(round_number)
          if (series_winner == team1) {
            permuted_df$rank_hat[permuted_df$true_rank == team2] <- 2
            permuted_df$rank_hat[permuted_df$true_rank == team1] <- 1
          } else {
            permuted_df$rank_hat[permuted_df$true_rank == team1] <- 2
            permuted_df$rank_hat[permuted_df$true_rank == team2] <- 1
          }
        }
      }

      teams <- next_round  # Winners move to the next round
      round_number <- round_number - 1
    }

    # Third place match between the losers of the semifinals
    if (third_place == TRUE){
      team1 <- losers_semis[1]
      team2 <- losers_semis[2]
      strength1 <- permuted_df$true_strength[permuted_df$true_rank == team1]
      strength2 <- permuted_df$true_strength[permuted_df$true_rank == team2]
      wins_team1 <- 0
      wins_team2 <- 0
      for (game in 1:series) {
        match_winner <- simulate_match(team1, team2, strength1, strength2)
        if (match_winner == team1) {
          wins_team1 <- wins_team1 + 1
        } else {
          wins_team2 <- wins_team2 + 1
        }
        # Determine the overall winner if a majority is reached
        if (wins_team1 > series / 2 || wins_team2 > series / 2) {
          break
        }
      }
      # Assign third place and fourth place ranks
      if (wins_team1 > wins_team2) {
        #cat(sprintf("Third Place Matchup: %s vs %s | Winner: %s\n", team1, team2, team1))
        permuted_df$rank_hat[permuted_df$true_rank == team1] <- 3
        permuted_df$rank_hat[permuted_df$true_rank == team2] <- 4
      } else {
        #cat(sprintf("Third Place Matchup: %s vs %s | Winner: %s\n", team1, team2, team2))
        permuted_df$rank_hat[permuted_df$true_rank == team2] <- 3
        permuted_df$rank_hat[permuted_df$true_rank == team1] <- 4
      }
    }

    permuted_df$seed <- rep(paste(seeding_structure[i, ], collapse = ", "), nrow(permuted_df))
    results[[i]] <- permuted_df # Store the results of this permutation
  }

  final_results <- do.call(rbind, results)

  # Remove ties if ties = FALSE
  if (!ties) {
    final_results <- final_results %>%
      filter(!is.na(true_rank)) %>%
      arrange(true_rank) %>%
      mutate(rank_hat = rank(rank_hat, ties.method = "random"))
  } else{
    final_results <- final_results %>%
      filter(!is.na(true_rank)) %>%
      arrange(rank_hat) %>%
      mutate(rank_hat = rank(rank_hat, ties.method = "average"))
  }

  final_results <- final_results %>%
    mutate(distribution = distribution) %>%
    mutate(true_rank = as.numeric(true_rank)) %>%
    arrange(true_rank)
  return(final_results)
}

knockout_teams <- simulate_groups(4,8,"Normal",2)
create_knockout_matches(knockout_teams)
```

>>>>

```{r}
simulate_group_stage_tournament <- function(num_teams, distribution, group_length = 4, ties = TRUE, rounds = 1, seeded=F) {

  if (num_teams <= 3) {
    stop("Number of Teams must be greater than 3.")
  }

  if (num_teams %% group_length != 0) {
    stop("Number of teams must be divisible by group length.")
  }

  teams <- paste(num_teams:1)

  # Create team strengths based on distribution
  if (distribution == "Normal"){
    strengths <- sapply(num_teams, function(n) { qnorm(1:n/(n+1)) })
    team_strengths <- data.frame(
      true_rank = as.numeric(teams),
      theta = strengths,
      rank_hat = rep(NA, length(teams)),
      game_wins = rep(0, length(teams)),
      game_losses = rep(0, length(teams))
    )
    df <- arrange(team_strengths, true_rank)
  }
  else if (distribution == "Same"){
    strengths <- sapply(num_teams, function(n) {rep(0, n)})
    team_strengths <- data.frame(
      true_rank = as.numeric(teams),
      theta = strengths,
      rank_hat = rep(NA, length(teams)),
      game_wins = rep(0, length(teams)),
      game_losses = rep(0, length(teams))
    )
    df <- arrange(team_strengths, true_rank)
  }
  else if (distribution == "Uniform"){
    strengths <- sapply(num_teams, function(n) { qunif(1:n/(n+1), 0, sqrt(12)) })
    team_strengths <- data.frame(
      true_rank = as.numeric(teams),
      theta = strengths,
      game_wins = rep(0, length(teams)),
      game_losses = rep(0, length(teams)),
      rank_hat = rep(NA, length(teams))
    )
    df <- arrange(team_strengths, true_rank)
  }
  else if (distribution == "Manual") {
    cat("Please enter the strengths for each team in descending rank order:\n")
    strengths <- numeric(num_teams)
    for (i in num_teams:1) {
      repeat {
        input <- as.numeric(readline(paste("Strength for Team", num_teams - i + 1, ": ")))
        if (!is.na(input)) {
          strengths[i] <- input
          break
        } else {
          cat("Invalid input. Please enter a numeric value.\n")
        }
      }
    }
    team_strengths <- data.frame(
      true_rank = as.numeric(teams),
      theta = strengths,
      rank_hat = rep(NA, length(teams)),
      game_wins = rep(0, length(teams)),
      game_losses = rep(0, length(teams))
    )
    df <- arrange(team_strengths, true_rank)
  }
  else {
    stop("Distribution not found: Enter Manual to input your own strengths")
  }

  num_groups <- num_teams / group_length

  # Create balanced groups using snake draft seeding
  # Sort teams by strength (best to worst)
  df <- df %>% arrange(desc(theta))

  # Assign groups using snake draft method
  df$groups <- NA
  for (i in 1:nrow(df)) {
    group_round <- ceiling(i / num_groups)
    if (group_round %% 2 == 1) {
      # Forward direction for odd rounds
      group_id <- ((i - 1) %% num_groups) + 1
    } else {
      # Reverse direction for even rounds
      group_id <- num_groups - ((i - 1) %% num_groups)
    }
    df$groups[i] <- group_id
  }

  # Reset game stats for group play
  df$game_wins <- 0
  df$game_losses <- 0
  df$rank_hat <- NA

  all_group_results <- list()

  # Simulate each group
  for (g in 1:num_groups) {
    group_teams <- df[df$groups == g, ]
    n_group_teams <- nrow(group_teams)

    # Play round robin within the group
    for (r in 1:rounds) {
      for (i in 1:(n_group_teams - 1)) {
        for (j in (i + 1):n_group_teams) {
          team1 <- group_teams$true_rank[i]
          team2 <- group_teams$true_rank[j]

          strength1 <- group_teams$theta[i]
          strength2 <- group_teams$theta[j]

          match_winner <- simulate_match(team1, team2, strength1, strength2)

          # Update wins and losses
          if (match_winner == team1) {
            df$game_wins[df$true_rank == team1] <- df$game_wins[df$true_rank == team1] + 1
            df$game_losses[df$true_rank == team2] <- df$game_losses[df$true_rank == team2] + 1
          } else {
            df$game_wins[df$true_rank == team2] <- df$game_wins[df$true_rank == team2] + 1
            df$game_losses[df$true_rank == team1] <- df$game_losses[df$true_rank == team1] + 1
          }
        }
      }
    }

    # Rank teams within each group
    group_df <- df[df$groups == g, ]
    if (ties) {
      group_df$group_rank <- rank(-group_df$game_wins, ties.method = "average")
    } else {
      group_df$group_rank <- rank(-group_df$game_wins, ties.method = "random")
    }

    all_group_results[[g]] <- group_df
  }

  # Combine all group results
  final_df <- do.call(rbind, all_group_results)

  # Calculate overall tournament ranking
  # First by group rank, then by total wins, then by true rank as tiebreaker
  final_df <- final_df %>%
    arrange(group_rank, desc(game_wins), true_rank)

  if (ties) {
    final_df$rank_hat <- rank(interaction(final_df$group_rank, -final_df$game_wins), ties.method = "average")
  } else {
    final_df$rank_hat <- rank(interaction(final_df$group_rank, -final_df$game_wins, final_df$true_rank), ties.method = "random")
  }

  final_df <- final_df %>%
    mutate(distribution = distribution) %>%
    arrange(true_rank) %>%
    mutate(rank_hat = ifelse(group_rank <= 2, group_rank, num_groups + group_rank)) %>%
    select(true_rank, theta, game_wins, game_losses, groups, group_rank, rank_hat, distribution)

  group_winners <- final_df %>%
    filter(group_rank == 1) %>%
    arrange(groups)

  group_seconds <- final_df %>%
    filter(group_rank == 2) %>%
    arrange(groups)

  n_groups <- length(unique(final_df$groups))
  matchup_vector <- c()

  # Build the alternating order
  for (i in seq(1, n_groups, by = 2)) {
    g1 <- group_winners$groups[i]
    g2 <- group_winners$groups[i + 1]

    winner_g1 <- group_winners$true_rank[group_winners$groups == g1]
    second_g2 <- group_seconds$true_rank[group_seconds$groups == g2]

    winner_g2 <- group_winners$true_rank[group_winners$groups == g2]
    second_g1 <- group_seconds$true_rank[group_seconds$groups == g1]

    # append in alternating order
    matchup_vector <- c(matchup_vector, winner_g1, second_g2, winner_g2, second_g1)
  }
  matchup_vector

  knockout_results <- simulate_knockout(df = final_df, matchup_vector = matchup_vector, series = 1, third_place = TRUE)

  knockout_results <- knockout_results %>%
    filter(!is.na(rank_hat))
  final_df <- final_df %>%
    filter(rank_hat > num_groups)

  combined_df <- bind_rows(
  final_df %>% filter(rank_hat >= num_groups),   # group-stage only teams
  knockout_results %>% filter(!is.na(rank_hat))  # knockout-stage teams
) %>%
  arrange(true_rank) %>%
    mutate(rank_hat = rank(rank_hat, ties.method = "random"))

  return(combined_df)
}

# Example usage:
# Single simulation
result <- simulate_group_stage_tournament(num_teams = 8, distribution = "Normal", group_length = 4, ties=F)

group_stage_results <- list()
for (teams in c(8)) {
  group_stage_results[[as.character(teams)]] <- list()
  for (replicate in 1:10000) {
    # Store each simulation result in the list with unique indexing
    group_stage_results[[as.character(teams)]][[replicate]] <- simulate_group_stage_tournament(
      num_teams = teams,
      distribution = "Normal",
      group_length = 4,
      ties = F
    )
  }
}

group_stage_results <- do.call(rbind, lapply(names(group_stage_results), function(teams) {
  team_results <- do.call(rbind, group_stage_results[[teams]])
  team_results$teams <- as.numeric(teams)
  return(team_results)
}))

group_stage_results %>% select(true_rank, game_wins, game_losses, groups, group_rank, rank_hat)

# Proportion of true #1 teams that finish group_rank > 2
sum(group_stage_results$true_rank == 1 & group_stage_results$group_rank > 2) /
sum(group_stage_results$true_rank == 1)

```

## Simulate Knockout Round

```{r}
simulate_knockout <- function(df, matchup_vector, series = 1, third_place = TRUE) {

  teams <- matchup_vector
  permuted_df <- df
  permuted_df$rank_hat <- NA

  round_number <- ceiling(log2(length(teams)))
  losers_semis <- c()

  while (length(teams) > 1) {

    #cat("\n--- New Round ---\n")
    next_round <- c()

    for (j in seq(1, length(teams), by = 2)) {
      team1 <- teams[j]
      team2 <- teams[j+1]

      #cat(sprintf("Matchup: Team1 = %s | Team2 = %s\n", team1, team2))

      strength1 <- permuted_df$theta[which(permuted_df$true_rank == team1)]
      strength2 <- permuted_df$theta[which(permuted_df$true_rank == team2)]

      #cat(sprintf("Strengths: %.2f vs %.2f\n", strength1, strength2))

      wins_team1 <- 0
      wins_team2 <- 0
      losses_team1 <- 0
      losses_team2 <- 0

      for (game in 1:series) {
        match_winner <- simulate_match(team1, team2, strength1, strength2)
        #cat(sprintf("Game %d Winner: %s\n", game, match_winner))

        if (match_winner == team1) {
          wins_team1 <- wins_team1 + 1
          losses_team2 <- losses_team2 + 1
        } else {
          wins_team2 <- wins_team2 + 1
          losses_team1 <- losses_team1 + 1
        }
        if (wins_team1 > series / 2 || wins_team2 > series / 2) break
      }

      # update df
      permuted_df <- permuted_df %>%
        mutate(game_wins = ifelse(true_rank == team1, game_wins + wins_team1, game_wins)) %>%
        mutate(game_wins = ifelse(true_rank == team2, game_wins + wins_team2, game_wins)) %>%
        mutate(game_losses = ifelse(true_rank == team1, game_losses + losses_team1, game_losses)) %>%
        mutate(game_losses = ifelse(true_rank == team2, game_losses + losses_team2, game_losses))

      # winner moves to next round
      series_winner <- ifelse(wins_team1 > wins_team2, team1, team2)
      #cat(sprintf("Series Winner: %s\n", series_winner))
      next_round <- c(next_round, series_winner)

      # assign ranks
      if (round_number > 2) {  # early rounds
        loser <- ifelse(series_winner == team1, team2, team1)
        permuted_df$rank_hat[permuted_df$true_rank == loser] <- 2^round_number
      } else if (round_number == 2) {  # semifinals
        loser <- ifelse(series_winner == team1, team2, team1)
        permuted_df$rank_hat[permuted_df$true_rank == loser] <- 4
        losers_semis <- c(losers_semis, loser)
      } else if (round_number == 1) {  # finals
        winner <- series_winner
        loser <- ifelse(series_winner == team1, team2, team1)
        permuted_df$rank_hat[permuted_df$true_rank == winner] <- 1
        permuted_df$rank_hat[permuted_df$true_rank == loser] <- 2
      }

    }

    teams <- next_round
    round_number <- round_number - 1
  }

  # third place match
  if (third_place & length(losers_semis) == 2) {
    team1 <- losers_semis[1]
    team2 <- losers_semis[2]
    strength1 <- permuted_df$theta[permuted_df$true_rank == team1]
    strength2 <- permuted_df$theta[permuted_df$true_rank == team2]

    #cat(sprintf("\nThird Place Match: %s vs %s | Strengths: %.2f vs %.2f\n", team1, team2, strength1, strength2))

    wins_team1 <- 0
    wins_team2 <- 0

    for (game in 1:series) {
      match_winner <- simulate_match(team1, team2, strength1, strength2)
      #cat(sprintf("Game %d Winner: %s\n", game, match_winner))

      if (match_winner == team1) wins_team1 <- wins_team1 + 1 else wins_team2 <- wins_team2 + 1
      if (wins_team1 > series / 2 || wins_team2 > series / 2) break
    }

    if (wins_team1 > wins_team2) {
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 3
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 4
    } else {
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 3
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 4
    }
  }

  permuted_df <- permuted_df %>% arrange(true_rank)
  return(permuted_df)
}

```
